---
title: "Penguin body mass by species and sex, in R"
author: "Abdulrahman Aleid"
date: "2026-09-23"
categories: [r, analysis]
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
collected by Dr. Kristen Gorman at Palmer Station Antarctica LTER and
released under CC0. Loaded from the `palmerpenguins` R package.
Citation: Horst AM, Hill AP, Gorman KB (2020). *palmerpenguins: Palmer
Archipelago (Antarctica) penguin data*. doi:10.5281/zenodo.3960218.

In this post I repeat my species comparison in R and add a second
question: does body mass also differ between male and female penguins
within each species?

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(knitr)
```

## Load and inspect the data

```{r}
penguins <- palmerpenguins::penguins
dim(penguins)
glimpse(penguins)
```

## Clean

```{r}
colSums(is.na(penguins))
```

```{r}
penguins_clean <- penguins |>
  filter(!is.na(body_mass_g), !is.na(flipper_length_mm))

nrow(penguins) - nrow(penguins_clean)
```

## Summarise

```{r}
#| label: tbl-mass-species
#| tbl-cap: "Body mass by species."
penguins_clean |>
  group_by(species) |>
  summarise(
    n = n(),
    mean_g = round(mean(body_mass_g), 1),
    sd_g = round(sd(body_mass_g), 1)
  ) |>
  arrange(desc(mean_g)) |>
  kable()
```

```{r}
#| label: tbl-mass-species-sex
#| tbl-cap: "Mean body mass by species and sex."
penguins_clean |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(n = n(), mean_g = round(mean(body_mass_g), 1), .groups = "drop") |>
  kable()
```

## Visualise

```{r}
#| label: fig-penguin-mass-r
#| fig-cap: "Body mass of penguins by species and sex."
penguins_clean |>
  filter(!is.na(sex)) |>
  ggplot(aes(x = species, y = body_mass_g, fill = sex)) +
  geom_boxplot() +
  labs(x = "Species", y = "Body mass (g)", fill = "Sex") +
  theme_minimal()
```

## What I found

@tbl-mass-species reproduces my Python result. Gentoo penguins are
by far the heaviest, averaging about 5,076 g, while Chinstrap and
Adelie are nearly identical at around 3,700 g. @tbl-mass-species-sex
and @fig-penguin-mass-r add the second question. In every species,
males are heavier than females, and the gap is visible in the boxplot
for all three. Sex is missing for a few birds, so those are left out
of the sex comparison.