---
title: "R and Python together: penguin body mass by species and sex"
author: "Abdulrahman Aleid"
date: "2026-09-23"
categories: [r, python, reticulate]
engine: knitr
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
collected by Dr. Kristen Gorman at Palmer Station Antarctica LTER and
released under CC0. Loaded from the `palmerpenguins` R package.
Citation: Horst AM, Hill AP, Gorman KB (2020). *palmerpenguins: Palmer
Archipelago (Antarctica) penguin data*. doi:10.5281/zenodo.3960218.

This post uses both languages in one document. R prepares the data,
Python calculates the average body mass for each species and sex, and
R takes that summary back to build a table and a chart. The two
languages share one session through the `reticulate` package.

## Setup (R)

```{r}
#| label: setup
#| message: false
#| warning: false
library(reticulate)
use_virtualenv("../../.venv", required = TRUE)

library(dplyr)
library(ggplot2)
library(knitr)
```

## Prepare the data (R)

```{r}
penguins_clean <- palmerpenguins::penguins |>
  filter(!is.na(body_mass_g), !is.na(sex)) |>
  select(species, sex, body_mass_g)

nrow(penguins_clean)
```

## Summarise (Python)

Python reads the R data frame as `r.penguins_clean`, which arrives as a
pandas DataFrame.

```{python}
import pandas as pd

penguins = r.penguins_clean
print(type(penguins))

penguins["body_mass_kg"] = penguins["body_mass_g"] / 1000

summary = (
    penguins
    .groupby(["species", "sex"], observed=True)["body_mass_kg"]
    .agg(n="count", mean_kg="mean")
    .round(2)
    .reset_index()
)
summary
```

## Bring the summary back (R)

R reads the Python DataFrame as `py$summary`, which arrives as an R
data frame.

```{r}
#| label: tbl-summary
#| tbl-cap: "Mean body mass (kg) by species and sex, computed in Python."
summary <- py$summary
class(summary)
kable(summary)
```

```{r}
#| label: fig-summary
#| fig-cap: "Mean body mass by species and sex, plotted in R from the Python summary."
ggplot(summary, aes(x = species, y = mean_kg, fill = sex)) +
  geom_col(position = "dodge") +
  labs(x = "Species", y = "Mean body mass (kg)", fill = "Sex") +
  theme_minimal()
```

## What I found

@tbl-summary and @fig-summary show that males are heavier than females
in all three species, and that Gentoo penguins are the heaviest for both
sexes. The averages were calculated in Python, and the table and chart
were made in R from that same result.